Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-25 young_pp_discs.tex
all changes from co-authors made

\date{Jan 2013}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionise this field, therefore it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines together allow for measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt*{Joos2012,Hayfield2011,Machida2011}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing the envelope material. \smallskip

After the pioneer work of, e.g., \citet{Chandler1995,Brown2000,Looney2000}, recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources (as defined by \citealt{Andre1999}), finding masses between 0.4 and  $>$1 M$_{\odot}$ \citep{Jorgensen2007,Jorgensen2009,Enoch2011}. A 130\,au disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ \citep{Choi2007}. \citet{Pineda2012} observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. \citet{Persson2012} observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets \citep{Hartogh2012}. \citet{Zapata2013} used ALMA to observe a disc of size $\sim$50$\,$au in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As fully-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \smallskip

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation \citep{Cossins2010,Ruge2013}. Molecular line emission from massive discs has been simulated by \citet{Krumholz2007}, assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of \citet{Boley2008}, where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., \citet{Visser2009,Visser2011}. As shown by \citet{Ilee2011}, hereafter I2011, the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of non-LTE 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation $\frac{31.5}{\sqrt{x+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT5.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.~\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to solving the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y, and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer is taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \citealt{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay triangulation means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (fig. \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. The average of ten separate runs was taken to minimise the artifacts in the output images, resulting from the grid construction (Fig. \ref{averages}).

@@

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical coordinates, linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\theta$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum2.eps}
 \caption{{\bf Left:} A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite gridding.}
 \label{averages}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way). For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the $^{17}$O/$^{16}$O elemental abundance. These estimates for the abundances in the envelope are simplistic, however, as seen in section \ref{sec:alma_predictions}, the envelope contribution to the line is very limited in velocity and can be spectrally disentangled from the disc contribution.\smallskip

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ relative to the edge on case.  In the following, we will show the 30$^\circ$ case, while the 15$^\circ$ is discussed in section \ref{sec:alma_predictions}. Other inclinations are shown in appendix \ref{sec:other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and  intensity weighted velocity maps (fig.s \ref{sim_all}, and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, which can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all2.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structure in different regions of the disc is highlighted by looking at different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are integrated over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (fig. \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}), it can be used to measure the rotation of the central part of the disc without contamination from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that they trace radii $<$16$\,$au, even smaller than the extent of the OCS column density distribution shown in fig.\ref{Chemistry}. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26$\rightarrow$25), as shown in fig. \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, this may be an artefact of our simplistic treatment of the inner hole, given that we neglect the presence of the protostar and other hot material within the central 2$\,$au. Thus, we expect absorption to be also observed toward the central region of the disc.\smallskip

@@

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/channel_map-2.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated by CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. Each map integrates over 2.4 km$\,$s$^{-1}$ with the top left from -10.8 to -8.4 km$\,$s$^{-1}$ to the bottom left ranging from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

>>>>

\date{Mar 2013}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs poorly constrained by observations. ALMA will soon revolutionise this field. Therefore, it is important to provide predictions which will be valuable for the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines can be used to probe the disc structure and kinematics at all scales. 
\end{abstract}

@@

Currently few observations constrain the formation and early evolution of protoplanetary discs around solar-type and low-mass protostars, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e. g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10 percent the original core mass (e. g.\citealt*{Joos2012,Hayfield2011,Machida2011}), are challenging. Sensitive interferometers are needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well as to filter out the extended emission tracing the envelope material. \smallskip

After the pioneering work of, e. g., \citet{Chandler1995,Brown2000,Looney2000}, recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources (as defined by \citealt{Andre1999}), finding masses between 0.4 and  $>$1 M$_{\odot}$ \citep{Jorgensen2007,Jorgensen2009,Enoch2011}. A 130\,au disc was discovered in NH$_3$ emission toward a Class 0 source in Perseus, with the Jansky Very Large Array (JVLA) \citep{Choi2007}. \citet{Pineda2012} observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. \citet{Persson2012} observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets \citep{Hartogh2012}. \citet{Zapata2013} used ALMA to observe a disc of size $\sim$50$\,$au in the other Class 0 source in IRAS 16293-2422 and detected inverse P-Cygni profiles in HCN and CH$_3$OH which indicate infall towards the disc. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As fully-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \smallskip

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation \citep{Cossins2010,Ruge2013}. Molecular line emission from massive discs has been simulated by \citet{Krumholz2007}, who assumed local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of \citet{Boley2008}, in which episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e. g. \citet{Visser2009,Visser2011}. As shown by \citet{Ilee2011}, hereafter I2011, the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies that retard the reaction at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of non-LTE 3D radiative transfer modelling, we have simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation$V_{rot}\frac{31.5}{\sqrt{R+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT5.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, in order to highlight the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.~\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to the solution of the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer are taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \citealt{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation and it is between the points connected by this method that photon are allowed to propagate (fig. \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 for 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. The average of ten separate runs was taken to minimise the artefacts in the output images, resulting from the grid construction (Fig. \ref{averages}).

@@

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their H$_2$ number density, and the number density of the molecular spieces being simulated, compared against those of a reference point in order to decide if the candidate point is to be used in the grid or not. Candidate grid points are selected at random in a cylindrical coordinates that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum2.eps}
 \caption{{\bf Left:} A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artefacts due to finite gridding.}
 \label{averages}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have reliable Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$. For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ at smaller radii \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the $^{17}$O/$^{16}$O isotope abundance. These estimates for the abundances in the envelope are simplistic. However, as seen in section \ref{sec:alma_predictions}, the envelope contribution to the line is very limited in velocity and can be spectrally disentangled from the disc contribution.\smallskip

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ relative to the edge on case. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and  intensity weighted velocity maps (figs. \ref{sim_all}, and \ref{mom0_maps}) were from -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid domination by the contribution from the envelope, which can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all2.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structures in different regions of the disc are highlighted by different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are obtained from integration over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (fig. \ref{sim_all}). As OCS is not seen in outflows (e. g. \citealt{Stanke2007}, \citealt{VDTak2003}), it can be used to measure the rotation of the central part of the disc without contamination from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that they trace radii $<$16$\,$au, even smaller than the extent of the OCS column density distribution shown in fig. \ref{Chemistry}. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26$\rightarrow$25), as shown in fig. \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, this may be an artefact of our treatment of the inner hole, given that we neglect the presence of the protostar and other hot material within the central 2$\,$au. Thus, we expect absorption also to be observed toward the central region of the disc.\smallskip

@@

\begin{figure*}
 \includegraphics[width=110mm]{Figures/sim/appendix_inclinations2.eps}
 \caption{C$^{17}$O 3$\rightarrow$2 {\bf left:} Continuum subtracted integrated intensity map (in K$\,$km$\,$s$^{-1}$). {\bf Centre:} Intensity weighted velocity map (in km$\,$s$^{-1}$). {\bf Right:} Position-velocity diagram with intensities in K. The inclinations for which results are shown from top to bottom are: 75$^\circ$, 45$^\circ$, 30$^\circ$ and 15$^\circ$ to edge on.}
 \label{other_inc}
\end{figure*}

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O. Fig. \ref{other_inc} shows the C$^{17}$O 3$\rightarrow$2 line in the same physical model but at 15$^\circ$, 30$^\circ$, 45$^\circ$ and 75$^\circ$ to edge on. At 15$^\circ$ to edge on we can still see some structure in the integrated intensity map and the structure is also visible in the position velocity diagram. At larger inclinations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/channel_map-2.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated with CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. the integration range for each map is 2.4 km$\,$s$^{-1}$, for the top left the range is from -10.8 to -8.4 km$\,$s$^{-1}$ and for the bottom right the range is from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

Fig. \ref{continuum} shows the 1\,mm continuum emission and beam size in the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where g is the gas/dust mass ratio S$_\nu$ is the integrated flux, d is the distance to the source, $\kappa_\nu$ is the dust opacity and B$_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. From this, a disc mass measurement of 0.1$\,$M$_\odot$ was obtained by assuming a dust temperature of 40K, and a measurement of 0.005$\,$M$_\odot$ was obtained by using 400K. The flux was integrated from y=0.2 to -0.2 and x=0.4 to -0.4. This shows that the disc is not optically thin to millimetre wavelengths. This underestimate of the disc mass by at least a factor of 4 shows that in young, massive discs such as those we are investigating, longer wavelength continuum observations are neccesary for determining the disc mass.\smallskip
